---
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[**Lab #2 Markdown File**](/assignments/Lab_2.Rmd)

[**Lab Instructions**](https://dssoc.github.io/assignments/lab_instructions.html)

In this lab we will practice using methods from the `tidyverse` package.

See the "Instructions" section of the [Introduction to Lab Assignments](https://dssoc.github.io/assignments/lab_instructions.html) page for more information about the labs. That page also gives descriptions for the datasets we will be using.

**Required reading:**

-   [R4DS Chapter 5: Data transformation](https://r4ds.had.co.nz/transform.html)
-   [R4DS Chapter 18: Pipes](https://r4ds.had.co.nz/pipes.html)
-   [R4DS Chapter 7: Exploratory data analysis](https://r4ds.had.co.nz/exploratory-data-analysis.html)

**Optional resources:**

-   [`gather` function documentation](https://tidyr.tidyverse.org/reference/gather.html#examples) (see examples)
-   [More dplyr examples](https://rstudio.cloud/learn/primers/2)
-   [More help with tidy](https://rstudio.cloud/learn/primers/4)
-   [R4DS Chapters 9-16](https://r4ds.had.co.nz/wrangle-intro.html)

```{r message=FALSE, echo=FALSE}
# Load the datasets and libraries. You shouldn't need to change the URL in the `load` function.

library(tidyverse)

load(url('https://dssoc.github.io/datasets/congress.RData'))
```

<br/>

## Example Questions

<br>

**ex1. how many female and male senators from the Democrat party are older than 60? For this and all assignments, you may approximate age as `approx age = current year - birthyear`.**

```{r}
# this solution uses pipes (see required reading)
congress %>% 
  filter(type=='sen', party=='Democrat', birthyear > 1962) %>% 
  count(gender)

# this is old-school
tmp <- filter(congress, type=='sen', party=='Democrat', birthyear > 1962)
count(tmp, gender)
```

**ex2. Give the names of the 2 oldest senators.**

```{r}
# this solution uses pipes
congress %>% 
  filter(type=='sen') %>% 
  arrange(birthyear) %>% 
  select(full_name) %>% 
  head(2)
```

**ex3. How many valid Twitter and Facebook account handles are there in `congress_contact`?**

```{r}
# your answer here
congress_contact %>% 
  select(facebook, youtube) %>% 
  gather(key='platform', value='handle', facebook, youtube) %>% 
  count(are_valid = handle!='') %>% 
  filter(are_valid)

congress_contact %>% 
  select(facebook, youtube) %>% 
  pivot_longer(facebook:youtube, names_to='platform', values_to='handle') %>% 
  count(are_valid = handle!='') %>% 
  filter(are_valid)
```

## Questions

<br>

**1. Describe what the following tidyverse functions do. Also describe the pipe operator "%\>%". You may need to look up the [official documentation](https://www.rdocumentation.org/) for each of these.**

```         
filter: used to subset certain rows of a dataset by setting a condition on a column (or "variable")
select: used to subset the dataset by certain columns
mutate: used to create a new column (or "variable") from a transformation of other column(s).
count: used to count the number of occurences of unique types in a column(s).
arrange: used to order rows of a dataset by column(s), can be ascending or descending.
gather: used to transform data from wide to long dataframe using key-value pairing
pivot_longer: used to transform dataframe from wide to long format, variables can be grouped using patterned matching, regular expression, transformations, etc
pipe operator ("%>%"): used to pass expressions (usually starting with a dataframe) through other expressions for data manipulation
```

<br/>

**2. How many male and female members are representatives and senators? Your output should appear as a single dataframe with four rows corresponding to female representatives, female senators, male representatives, and male senators.**

NOTE: you can identify representators by `type=="rep"` and senators using `type=="sen"`.

```{r}
# your answer here
congress %>% 
  count(gender,type)
```

<br/>

**3. Create a dataframe showing the oldest and youngest female senator democrats using only R code.**

HINT: check out the `slice()` function.

```{r}
# your answer here
oldest_youngest <- congress %>% 
  filter(gender=="F",
         party == "Democrat",
         type == "sen") %>% 
  arrange(birthdate) %>% 
  slice(1, n())
oldest_youngest
```

<br/>

**4. Using `mutate`, create a new variable called `age` which represents the approximate age of each member of congress. How many democratic senators are over 60 years old?**

Note: For this and all future questions this semester, you can approximate age using the formula `age = 2022-birthyear` (or replace 2022 with the current year).

```{r}
# your answer here
congress <- congress %>% 
  mutate(age = 2024 - birthyear)

congress %>% 
  filter(type == 'sen',
         party == 'Democrat',
         age >60) %>% 
  summarise(n())
```

<br/>

**5. Create a new column that indicates whether or not the member of congress is more than 55 years old, and create a single dataframe showing the number of male and female members of congress that are over and under 55.**

Note: the dataframe should have four rows: number of females over 55, number of males over 55, number of females 55 and under, number of males 55 and under.

```{r}
# your answer here
congress <- congress %>% 
  mutate(more_than_55 = case_when(age > 55 ~ TRUE,
                                  TRUE ~ FALSE))
more_than_55_df <- congress %>% 
  count(gender,more_than_55)
more_than_55_df
```

<br/>

**6. For this problem, use the `congress_contact` dataframe that includes the contact information for each member of congress. Using `gather`, create a new dataframe where each row corresponds to a valid twitter, facebook, or youtube social media account username (for youtube you can ignore `youtube_id`), and compute *the total number of valid accounts across all congress members*. Then accomplish the same task using `pivot_longer`.**

HINT: see the link to the `gather` function documentation in the recommended readings to get a better sense of how it works. You can use the example data they show to try it for yourself.

HINT: the account name will be an empty string in cases where the member of congress does not have a valid user account, so you can filter using the conditional `column_name != ''`. This compares each value in the column called `column_name` to an empty string to determine whether it is a valid account.

```{r}
# your answer here
sm_gathered <- gather(congress_contact, key = social_media_type, value = account_name, twitter:youtube)

sm_gathered %>% 
  filter(account_name != '') %>% 
  summarise(n())

sm_pivlonger <- pivot_longer(congress_contact, cols = twitter:youtube,names_to = "social_media_type",values_to = "account_name")

sm_pivlonger %>% 
  filter(account_name != '') %>% 
  summarise(n())
```

<br/>

**Below, I define for you two vectors corresponding to policies that US States have adopted to respond to COVID-19: restrictions on travel ([recorded May 20, 2020](https://www.wsj.com/articles/a-state-by-state-guide-to-coronavirus-lockdowns-11584749351)) and requirements that citizens to wear masks in public ([recorded August 17, 2020](https://www.cnn.com/2020/06/19/us/states-face-mask-coronavirus-trnd/index.html)).**

```{r}
travel_restrictions <- c("WA", "OR", "NV", "CA", "NM", "MN", "IL", "OH", "MI", "PA", "VA", "NY", "MA", "VH", "ME", "DE", "MD", "NJ")

require_masks <- c("HI", "WA", "OR", "NV", "CA", "MT", "CO", "NM", "KS", "TX", "MN", "AR", "LA", "WI", "IL", "AL", "MI", "IN", "OH", "KY", "WV", "NC", "VA", "DC", "DE", "PA", "NY", "VT", "NH", "MA", "RI", "CT", "ME")
```

<br/>

**7. write code to print only the states who implemented both travel restrictions and mask requirements:**

```{r}
# your answer here
intersect(travel_restrictions,require_masks)
```

<br/>

**8. Write code to print the states who had implemented mask requirements but not travel restrictions:**

```{r}
# your answer here
setdiff(require_masks, travel_restrictions)
```

**9. Describe two broad topics you might be interested in exploring for your final project. How would you use data science to gain insight about these topics? We won't require you to stick with these topics - we just want to see you brainstorming about what you might be interested in.**

```         
1) I am interested in finding a way to quantify the time period and frequency of trending social issues on a specific media platform. I can use data science by accumulating data on recent social issue postings by way of hashtags, visualize the frequency of those hashtags over units of time, and I can do some hypothesis testing on whether there is a statistically significant difference on the time period/frequency of multiple social issues.

2) I am also interested in using geographic data in some way relating transportation. I can use data science by accumulating data of surveys or from city transportation companies, creating mapped visualizations of different transportation modes and other demographic characteristics, and possibly create a regression with predictors (including location) that may be associated with modes of transportation.
```
